---
title: "scRNA-seq analysis for mouse heart samples"
output: html_document
date: "2025-04-25"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
options(future.globals.maxSize = 3*1024^3)   # 3 GiB cap: enough for SCTransform, but a SAFETY BRAKE
                                             # so a runaway step errors instead of exhausting RAM. Do NOT set to Inf.
library(Seurat)
library(SeuratObject)
# packageVersion("Seurat")
library(dplyr)
library(tibble)
library(patchwork)
library(dittoSeq)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)        # treeplot, pairwise_termsim
library(EnhancedVolcano)   # volcano plot
library(DT)                # datatable()
# MAESTRO is optional (cell-type annotation only); guarded so this runs without it.
have.maestro <- requireNamespace("MAESTRO", quietly = TRUE)
if (have.maestro) suppressMessages(library(MAESTRO))

## parameters
timestage <- "P0"
# Output root for derived files (processing/ + results/). The runner sets
# SCRNA_OUT_DIR="." so outputs land in our_analysis/. Defaults to ".".
out.dir <- Sys.getenv("SCRNA_OUT_DIR", unset = ".")
# Portable working directory: when knitting, this chunk runs in 03_condition_merge/,
# so ".." is the our_analysis root that holds processing/ and results/.
knitr::opts_knit$set(root.dir = normalizePath(".."))

seurat.KO <- readRDS(file.path(out.dir, "processing", paste0("merge.lanes.", timestage, "KO.rds")))
seurat.WT <- readRDS(file.path(out.dir, "processing", paste0("merge.lanes.", timestage, "WT.rds")))

seurat.KO$orig.ident <- paste0(timestage, "KO")
seurat.WT$orig.ident <- paste0(timestage, "WT")

seurat.KO
seurat.WT
```

## Merging seurat obj

```{r, fig.width=12, fig.height=6}
merge.scRNA <- merge(x = seurat.KO, y=seurat.WT, add.cell.ids = c("KO", "WT"), project = "Han_scRNA")
# v5: rejoin the per-sample count layers created by merge() (no-op on v4).
if (utils::packageVersion("SeuratObject") >= "5.0.0") merge.scRNA <- JoinLayers(merge.scRNA)
merge.scRNA$orig.ident <- factor(merge.scRNA$orig.ident)

head(merge.scRNA@meta.data)

VlnPlot(merge.scRNA, features = "nFeature_RNA", split.by = "orig.ident")
VlnPlot(merge.scRNA, features = "nCount_RNA", split.by = "orig.ident")
rm(seurat.KO, seurat.WT)
```

```{r}
merge.scRNA <- SCTransform(merge.scRNA, method = "glmGamPoi", verbose = F, conserve.memory = T)


# resolution = 0.8
merge.scRNA <- RunPCA(merge.scRNA, verbose = F) %>% 
  FindNeighbors(dims= 1:30) %>%
  FindClusters() %>% 
  RunUMAP(reduction = "pca", dims = 1:30) 


```



```{r, fig.width=12, fig.height=6}
DimPlot(merge.scRNA, group.by = "orig.ident")
DimPlot(merge.scRNA, group.by = "seurat_clusters")
# Lane-effect / batch check: if clusters separate by lane here, the lanes need
# integration (e.g. Harmony) rather than a plain merge() -- see README phase 2.
if (!is.null(merge.scRNA$lane)) DimPlot(merge.scRNA, group.by = "lane")
dittoDimPlot(merge.scRNA,var="orig.ident", split.by = "orig.ident")
dittoDimPlot(merge.scRNA,var="seurat_clusters", split.by = "orig.ident")
```

## DEG

```{r, fig.width=12, fig.height=6}
merge.scRNA <- PrepSCTFindMarkers(merge.scRNA)
merge.scRNA <- SetIdent(merge.scRNA, value = "seurat_clusters")

all.markers <- FindAllMarkers(merge.scRNA, verbose = F)
all.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

top30 <- all.markers %>% group_by(cluster) %>% top_n(n = 30, wt = avg_log2FC)
DoHeatmap(merge.scRNA, features = top30$gene) + NoLegend()
```

## Cell type annotation


```{r, fig.width=12, fig.height=6}
if (have.maestro) {
  data(mouse.all.droplet.TabulaMuris)
  sig.mouse.heart <- mouse.all.droplet.TabulaMuris[grep("\\(Heart_and_Aorta\\)", mouse.all.droplet.TabulaMuris$cluster) ,]
  sig.mouse.heart$cluster <- gsub("\\(Heart_and_Aorta\\)", "", sig.mouse.heart$cluster)
  anno.markers <- all.markers
  rownames(anno.markers) <- NULL
  colnames(anno.markers) <- c("p_val","avg_logFC","pct.1","pct.2","p_val_adj","cluster","gene")

  merge.scRNA <- RNAAnnotateCelltype(RNA= merge.scRNA, genes = anno.markers,
                                      signatures = sig.mouse.heart)

  print(table(merge.scRNA$assign.ident))
  DimPlot(merge.scRNA, group.by = "assign.ident", label = T)
} else {
  message("MAESTRO not installed -- skipping annotation; cardiomyocytes will be selected by markers below.")
}
```

## Cell cycle scoring

```{r}
# Cell-cycle gene lists from Seurat's built-in human set, mapped to mouse
# orthologs. Prefer a real ortholog table (babelgene); fall back to title-casing
# only if the package isn't installed. This removes the dependency on the
# external Mus_musculus.csv that the original analysis used.
to_mouse <- function(genes) {
  if (requireNamespace("babelgene", quietly = TRUE)) {
    o <- babelgene::orthologs(genes = genes, species = "mouse", human = TRUE)
    unique(o$symbol)
  } else {
    message("Package 'babelgene' not installed; using title-case approximation for mouse symbols.")
    paste0(substr(genes, 1, 1), tolower(substr(genes, 2, nchar(genes))))
  }
}
s.genes   <- sort(to_mouse(cc.genes.updated.2019$s.genes))
g2m.genes <- sort(to_mouse(cc.genes.updated.2019$g2m.genes))

# --- Original approach (re-enable if you have the Mus_musculus.csv file) ---
# phase.genes <- read.csv("scripts/Mus_musculus.csv")
# s.genesID <- phase.genes %>% filter(phase == "S") %>% pull(geneID)
# s.genes   <- bitr(s.genesID, fromType = "ENSEMBL", toType = "SYMBOL", OrgDb = org.Mm.eg.db) %>% pull(SYMBOL) %>% sort()
# g2m.genesID <- phase.genes %>% filter(phase == "G2/M") %>% pull(geneID)
# g2m.genes   <- bitr(g2m.genesID, fromType = "ENSEMBL", toType = "SYMBOL", OrgDb = org.Mm.eg.db) %>% pull(SYMBOL) %>% sort()
```



```{r,  fig.width=12, fig.height=10}
merge.scRNA <- CellCycleScoring(merge.scRNA, s.features = s.genes, 
                                      g2m.features = g2m.genes, set.ident = TRUE)

merge.scRNA <- SetIdent(merge.scRNA, value="seurat_clusters")
RidgePlot(merge.scRNA, features = c("S.Score", "G2M.Score"), ncol = 2)
```

## Subset cardiac cells

Cardiomyocytes are selected by **cell-type annotation / canonical markers** so
the selection is reproducible. (The original code drew a fixed circle on the
UMAP — within radius 9 of point (4,0) — which silently selects the wrong cells
once re-clustering shifts the embedding.)

```{r, fig.width=12, fig.height=8}
# Primary: cells annotated as cardiomyocyte/cardiac muscle by RNAAnnotateCelltype.
cm.pattern <- "cardiac|myocyte|cardiomyocyte"
# Seurat v5 errors on $ access to a missing meta column (v4 returned NULL), and
# assign.ident only exists when MAESTRO ran. Guard it so we fall through to the
# marker-score fallback below when annotation is absent (the MAESTRO-skip path).
is.cardiac <- if ("assign.ident" %in% colnames(merge.scRNA@meta.data)) {
  grepl(cm.pattern, merge.scRNA$assign.ident, ignore.case = TRUE)
} else {
  rep(FALSE, ncol(merge.scRNA))
}

# Fallback: if annotation labels don't match, score canonical CM markers and
# keep the top-scoring clusters.
cm.markers <- intersect(c("Tnnt2", "Myh6", "Actc1", "Nppa", "Myh7", "Ttn"),
                        rownames(merge.scRNA))
if (sum(is.cardiac) < 50) {
  message("Annotation-based cardiomyocyte selection sparse; using marker score fallback.")
  merge.scRNA <- AddModuleScore(merge.scRNA, features = list(cm.markers), name = "CMscore")
  cl.score <- tapply(merge.scRNA$CMscore1, merge.scRNA$seurat_clusters, median)
  cm.clusters <- names(cl.score)[cl.score > quantile(cl.score, 0.75, na.rm = TRUE)]
  is.cardiac <- merge.scRNA$seurat_clusters %in% cm.clusters
}
cells.select <- colnames(merge.scRNA)[is.cardiac]
cardiac.cells <- merge.scRNA[, cells.select]
cardiac.cells <- SetIdent(cardiac.cells, value = "orig.ident")

cat(sprintf("Selected %d cardiomyocytes (%.1f%% of cells)\n",
            length(cells.select), 100 * length(cells.select) / ncol(merge.scRNA)))

# EXPLORATORY ONLY: single-cell Wilcoxon. This treats cells as independent
# replicates (pseudoreplication) and is NOT a valid condition-level test --
# use the pseudobulk results below for inference. See the pseudobulk chunk.
cardiac.cells <- PrepSCTFindMarkers(cardiac.cells)
markers.2cond <- FindMarkers(cardiac.cells, ident.1 = "P0KO", ident.2 = "P0WT")


DimPlot(merge.scRNA, cells.highlight = cells.select)
DT::datatable(markers.2cond)
EnhancedVolcano(markers.2cond,
                lab = rownames(markers.2cond),
                x = 'avg_log2FC',
                y = 'p_val_adj',
                ylim = c(0, 100),
                xlab = bquote(~Log[2]~ 'fold change'),
                ylab = bquote(~-Log[10]~ 'FDR'),
                pCutoff = 0.01,
                FCcutoff = 2,
                pointSize = 2.0,
                labSize = 3.0,
                labCol = 'black',
                labFace = 'bold',
                boxedLabels = FALSE,
                colAlpha = 0.5,
                legendPosition = 'right',
                legendLabSize = 10,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 0.2,
                colConnectors = 'black')  +  
  theme(axis.text.x=element_text(size=10,face="bold"), axis.text.y=element_text(size=10,face="bold"), axis.title=element_text(size=12,face="bold"))
```

## Pseudobulk DE (recommended for KO vs WT)

```{r pseudobulk-DE, fig.width=10, fig.height=8}
## Aggregate raw counts per replicate, then test with DESeq2. This is the
## statistically appropriate way to compare conditions.
## VALID ONLY if lane1 and lane6 are DIFFERENT animals (biological replicates).
## If both lanes are the same animal, n=1 per condition and these p-values are
## not interpretable -- confirm the replicate structure with the lab first.
suppressMessages(library(DESeq2))

if (!is.null(cardiac.cells$lane)) {
  cardiac.cells$pb.sample <- paste(cardiac.cells$orig.ident, cardiac.cells$lane, sep = "_")
  pb <- AggregateExpression(cardiac.cells, assays = "RNA", group.by = "pb.sample",
                            slot = "counts")$RNA
  # AggregateExpression (Seurat v5) rewrites "_" in group names to "-", so strip a
  # "-lane#" OR "_lane#" suffix to recover the condition (e.g. P0WT-lane1 -> P0WT).
  coldata <- data.frame(sample = colnames(pb),
                        condition = sub("[-_]lane[0-9]+$", "", colnames(pb)))
  rownames(coldata) <- coldata$sample   # DESeq2 requires colData rows to match count cols
  coldata$condition <- relevel(factor(coldata$condition), ref = paste0(timestage, "WT"))

  if (length(unique(coldata$condition)) == 2 && all(table(coldata$condition) >= 2)) {
    dds <- DESeqDataSetFromMatrix(round(as.matrix(pb)), colData = coldata, design = ~ condition)
    dds <- dds[rowSums(counts(dds)) >= 10, ]
    dds <- DESeq(dds)
    res.pb <- as.data.frame(results(dds))
    res.pb <- res.pb[order(res.pb$padj), ]
    write.csv(res.pb, file = file.path(out.dir, "results/markers", paste0(timestage, ".cardiac.pseudobulk.DESeq2.csv")))
    DT::datatable(head(res.pb, 200))
    EnhancedVolcano(res.pb, lab = rownames(res.pb),
                    x = "log2FoldChange", y = "padj",
                    pCutoff = 0.05, FCcutoff = 1,
                    title = paste0(timestage, " KO vs WT (cardiac, pseudobulk)"))
  } else {
    message("Pseudobulk needs >= 2 replicates per condition; use the exploratory Wilcoxon above only.")
  }
} else {
  message("No `lane` metadata -- re-run the updated scRNA_lane_merge.Rmd to enable pseudobulk DE.")
}
```


```{r, fig.width=12, fig.height=6}
pos.markers <- markers.2cond %>% filter(avg_log2FC > 1, p_val < 0.01) %>% rownames()
GO.res <- enrichGO(pos.markers, OrgDb = "org.Mm.eg.db", keyType = "SYMBOL", ont = "BP",
                   pvalueCutoff = 0.05, qvalueCutoff = 0.2, readable=T)
# This GO step is exploratory (built on the pseudoreplication Wilcoxon markers) and
# sits downstream of the critical saveRDS below. enrichplot's pairwise_termsim/
# treeplot/barplot error when no terms are *significantly* enriched (note:
# nrow(GO.res@result) counts ALL tested terms, not the significant ones -- use
# as.data.frame() for the significant count). Make the whole block non-fatal so the
# merged object is always saved.
n.sig <- tryCatch(nrow(as.data.frame(GO.res)), error = function(e) 0L)
if (!is.null(GO.res) && isTRUE(n.sig >= 1)) {
  tryCatch({
    # Persist the GO table -- in batch mode the datatable widget below is not captured.
    write.csv(as.data.frame(GO.res), file = file.path(out.dir, "results/markers",
              paste0(timestage, ".cardiac.exploratory.GO_BP.csv")))
    GO.res <- pairwise_termsim(GO.res, method = "JC")
    if (n.sig >= 5) print(treeplot(GO.res))
    print(mutate(GO.res, qscore = -log(p.adjust, base=10)) %>% barplot(x="qscore"))
    datatable(GO.res@result, extensions="Buttons",
              options = list(dom="Blfrtip", buttons=c("copy","csv","excel","pdf","print")))
  }, error = function(e) message("GO plotting skipped: ", conditionMessage(e)))
} else {
  message(sprintf("enrichGO found %s significantly enriched GO BP term(s); skipping GO plots/CSV.",
                  ifelse(is.na(n.sig), 0L, n.sig)))
}
```


```{r save-merged, eval=TRUE}
saveRDS(merge.scRNA, file = file.path(out.dir, "processing", paste0("seurat.", timestage, ".merge.rds")))
# merge.scRNA <- readRDS(file.path(out.dir, "processing", "seurat.P0.merge.rds"))
```

